---
title: "Lab 5 - Computational Statistics (732A89)"
author: "Helena Llorens Lluís (hllor282), Yi Yang (yiyan338)"
date: "2025-02-20"
output: pdf_document
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

```{r, echo=FALSE}
library(ggplot2)
library(knitr)
library(boot)
```

# QUESTION 1 Bootstrap for regression

First, we use `lm()` to fit a cubic regression model  $$y=\beta_0+\beta_1x+\beta_2x^2+\beta_3x^3+\epsilon$$,where $x$ represnts the *concentration* of the fertilizer(%) and $y$ represents the *yield*(mg).

```{r echo=FALSE}
data <- read.csv("kresseertrag.dat",header=FALSE,sep = "")
colnames(data) <- c("Number","Fertilizer","Yield")
data <- as.data.frame(data)

#a
modelA <- lm(Yield~poly(Fertilizer,3,raw = TRUE),data = data)
summary(modelA)
```
To reduce model complexity, we remove the cubic term and fit a *quardic model* $$y=\beta_0+\beta_1x+\beta_2x^2$$ We  then  estimate the coefficients  with their 95% confidence intervals. The regression curve plot and confidence intervals for the coefficients are shown below:

```{r, fig.cap='Yields vs Concentration', echo=FALSE, fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
#b
#remove a term
modelB <- lm(Yield~poly(Fertilizer,2,raw = TRUE),data = data)
#plot
ggplot(data, aes(x = Fertilizer, y = Yield)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), color = "#4E79A7") +
  labs( x = "Concentration", y = "Yields")


#coefficients with 95% confidence interval
CI <- confint(modelB,level = 0.95)
kable(CI,caption = "95% Confidence Interval of parameters")
```

We now use the **bootstrap method** with 10,000 replicates to derive a 95% confidence interval for **$\beta_1$** using percentile method.The histogram and the 95% confidence interval are shown below:

```{r echo=FALSE ,fig.cap='Bootstrap Coefficient Distribution',fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
#c bootstrap for beta1
bo <- 10000                         #bootstrasp replicates
bs <- numeric(bo)
set.seed(12345)
#save the results
for (i in 1:bo) {
  #sampling using indcies
  indices <- sample(1:nrow(data),size=nrow(data),replace = TRUE)
  bootstrapData <- data[indices,]
  model_bootstrap <- lm(Yield~poly(Fertilizer,2,raw = TRUE),data = bootstrapData)
  bs[i] <- coef(model_bootstrap)[2]
}

bs_data <- data.frame(bs)
#hist(bs)
ggplot(bs_data,aes(x=bs))+
  geom_histogram(binwidth = 10,fill="#4E79A7",color="blue")+
  theme_minimal()+
  labs(x="Beta1",y="Frequency")
bss <- sort(bs)
ci95 <- unname(c(bss[round(bo*0.025)],bss[round(bo*0.975)]))
cat("95% CI for beta 1 :",ci95,"\n")


```

Now we use the package `boot`  to compute a 95% confidence interval for $\beta_1$ using both percentile method and BCa(bias-corrected and accelerated) method.The table below summarizes the 95% confidence intervals obtained using different methods:

```{r echo=FALSE}
#d bootstrap using boot package
beta1 <- function(data,i){
  model <- lm(Yield~poly(Fertilizer,2,raw = TRUE),subset=i,data = data)
  coef(model)[2]
}

cb <- boot(data,beta1,R=10000)
perc <- boot.ci(cb,type = "perc")
bca <- boot.ci(cb,type = "bca")
#cat("95% CI with percentile method :","\n")
#perc

#cat("95% CI with percentile method: ", perc$percent[4],  perc$percent[5], "\n")
#cat("95% CI with bca method: ", bca$bca[4], bca$bca[5], "\n")

```



```{r echo=FALSE}
 result <- c(CI[2,],95.3744,ci95,87.0976,perc$percent[4],perc$percent[5],87.3948,bca$bca[4],bca$bca[5],87.7167)
 result_mt <- matrix(result,byrow = TRUE,ncol=3)
 rownames(result_mt) <- c("lm","bootstrap","boot_percentile","boot_BCa")
 colnames(result_mt) <- c("Lower Bound","Upper Bound","Interval Width") 
 kable(result_mt,caption = "95% Confidence Interval of different methods")

```

**Discussion of Confidence Interval Differences**

From the results, we observe the following:

1.**The confidence interval(CI) using `lm()`is wider than the bootstrap CIs.** This is because `lm()` assumes that data  follows a normal distribution.If the data sample size is small or the data is not normally distributed this assumption may not hold, leading to a less accurate CI.

2.**The bootstrap method produces narrower confidence interval without distributional assumption.**  This allows it to better adapt to the actual data distribution.

3.**The CIs obtained using the manual bootstrap method and the `boot` package  with the `percentile` method are similar.**   However, the CI from the `boot` package with the `BCa` method is slightly different. This difference may be due to the BCa method correcting for bias, suggesting that the data may have some skewness.


# Appendix
## Question 1
```{r eval=FALSE}
library(ggplot2)
library(boot)
data <- read.csv("kresseertrag.dat",header=FALSE,sep = "")
colnames(data) <- c("Number","Fertilizer","Yield")
data <- as.data.frame(data)

#a
modelA <- lm(Yield~poly(Fertilizer,3,raw = TRUE),data = data)
summary(modelA)

confint(modelA,level = 0.95)


#b
#remove a term
modelB <- lm(Yield~poly(Fertilizer,2,raw = TRUE),data = data)
summary(modelB)

#coefficients with 95% confidence interval
confint(modelB,level = 0.95)

#plot
#plot(data$Fertilizer,data$Yield,main="Yields vs Concentration",
 #    xlab = "Concentration",ylab = "Yields")
#lines(data$Fertilizer,fitted(modelB),col="blue")

ggplot(data,aes(x=Fertilizer,y=Yield))+
  geom_point()+
  labs(title = "Yields vs Concentration",x="Concentration",y="Yields")+
  geom_line(aes(y=fitted(modelB)),color="#4E79A7")
  #geom_line(aes(y=fitted(modelA)),color="red")

#c bootstrap for beta
bo <- 10000                         #bootstrasp replicates
bs <- c()
set.seed(12345)
#save the results
for (i in 1:bo) {
  #sampling using indcies
  indices <- sample(1:nrow(data),size=nrow(data),replace = TRUE)
  bootstrapData <- data[indices,]
  model_bootstrap <- lm(Yield~poly(Fertilizer,2,raw = TRUE),data = bootstrapData)
  bs <- c(bs,coef(model_bootstrap)[2])
}
hist(bs)
bss <- sort(bs)
ci95 <- c(bss[round(bo*0.25)],bss[round(bo*0.975)])
ci95


#d bootstrap using boot package
beta1 <- function(data,i){
  model <- lm(Yield~poly(Fertilizer,2,raw = TRUE),subset=i,data = data)
  coef(model)[2]
}

cb <- boot(data,beta1,R=10000)
perc <- boot.ci(cb,type = "perc")
bca <- boot.ci(cb,type = "bca")
perc

 result <- c(CI[2,],95.3744,ci95,87.0976,perc$percent[4],perc$percent[5],87.3948,bca$bca[4],bca$bca[5],87.7167)
 result_mt <- matrix(result,byrow = TRUE,ncol=3)
 rownames(result_mt) <- c("lm","bootstrap","boot_percentile","boot_BCa")
 colnames(result_mt) <- c("Lower Bound","Upper Bound","Interval Width") 
 kable(result_mt,caption = "95% Confidence Interval of different methods")
```
