---
title: "Lab 1 - Computational Statistics (732A89)"
author: "Helena Llorens Lluís (hllor282), Yi Yang (yiyan338)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(ggplot2)
```


# QUESTION 1

First, we plot the function $$f(x, y) = sin(x + y) + (x - y)^2 - 1.5x + 2.5y + 1$$ within the intervals $x \in [-1.5, 4]$ and $y \in [-3, 4]$.

```{r, echo=FALSE, fig.dim=c(4, 2.5), fig.align='center'}
# define the function f(x, y) that returns f(x, y) if both x and y are in the interval
f <- function(x, y){
  ifelse(x >= -1.5 & x <= 4 & y >= -3 & y <= 4,
         sin(x + y) + (x - y)^2 - 1.5*x + 2.5*y + 1,
         NA)
}

# data frame with x, y and f(x, y)
x <- seq(-1.5, 4, length.out = 100)
y <- seq(-3, 4, length.out = 100)

grid <- expand.grid(x = x, y = y)
grid$z <- with(grid, f(x, y))

# plot f(x, y)
ggplot(grid, aes(x = x, y = y, z = z)) + 
  geom_contour(col = "#4E79A7") + 
  labs(x = "x", y = "y", title = "Contour plot of f(x, y)") + 
  theme_minimal()
```


```{r, echo=FALSE}
# gradient for f(x, y)
gradient <- function(x, y){
  grad_x <- cos(x + y) + 2*(x - y) - 1.5
  grad_y <- cos(x + y) - 2*(x - y) + 2.5
  return(c(grad_x, grad_y))
}

# hessian matrix for f(x, y)
hessian <- function(x, y){
  
  m <- matrix(c(-sin(x + y) + 2, -sin(x + y) - 2,
              -sin(x + y) - 2, -sin(x + y) + 2),
              byrow = T, nrow = 2, ncol = 2)
  
  return(m)
}
```

```{r, echo=FALSE}
# newton algorithm
newton <- function(gradient, hessian, x_init, y_init, tol = 0.0001, maxit = 100){
  
  # set initial values
  x0 <- c(x_init, y_init)
  
  # while the stopping criteria is not met
  for(it in 1:maxit){
    
    # add regularization so that the inverse can be calculated
    if(det(hessian(x0[1], x0[2])) == 0){
      x1 <- x0 - as.vector(solve(hessian(x0[1]+1e-12, x0[2]+1e-12)) %*% gradient(x0[1], x0[2]))
    } else {
      # update values
      x1 <- x0 - as.vector(solve(hessian(x0[1], x0[2])) %*% gradient(x0[1], x0[2]))
    }
    
    x1[1] <- max(min(x1[1], 4), -1.5)
    x1[2] <- max(min(x1[2], 4), -3)
    
    # if the stopping criteria is met
    if(norm(x1 - x0, type = "2") < tol){  # t(x1 - x0) %*% (x1 - x0)
      return(list(x = x1[1], y = x1[2], iterations = it, 
               gradient = gradient(x1[1], x1[2]), hessian = hessian(x1[1], x1[2])))
    } 
    
    x0 <- x1
  }
  
  warning("Maximum number of iterations reached without convergence.")
  return(list(x = x1[1], y = x1[2], iterations = maxit,
           gradient = gradient(x1[1], x1[2]), hessian = hessian(x1[1], x1[2])))
}
```

With both the gradient and the Hessian matrix calculated, we have been able to compute the Newton algorithm to find local minimums of this function. This function takes as input both the gradient and the Hessian matrix, a pair of starting values $x_0$ and $y_0$, and a tolerance (set at default as $0.0001$) and a maximum of iterations (set at $100$ as default).



```{r}
x_init <- 1
y_init <- 3
newton(gradient, hessian, x_init, y_init)
```


```{r}
x_init <- 2
y_init <- 4
newton(gradient, hessian, x_init, y_init)
```


```{r}
x_init <- 0
y_init <- 0
newton(gradient, hessian, x_init, y_init)
```

```{r}
x_init <- 0
y_init <- 1
newton(gradient, hessian, x_init, y_init)
```

