---
title: "Lab 3 - Computational Statistics (732A89)"
author: "Helena Llorens Lluís (hllor282), Yi Yang (yiyan338)"
date: "2025-02-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE,warning=FALSE}
library(ggplot2)
library(knitr)
suppressWarnings(library(knitr))
```

```{r include=FALSE}
#question1
#a
library(ggplot2)
fx <- function(x){
  if(x< -1|x > 1){
    return(0)}
  if(-1 <= x&& x <= 0){
    return(x+1)}
  if(0 < x && x <= 1){
    return(1-x)}
}

rejection_sample <- function(n){
  samples <- numeric(n)
  count <- 0
  
  while (count<n) {
    sample <- runif(1,-1,1) #sampling from uniform distribution
    U <- runif(1)
    f_x <- fx(sample)
    
    if(U <= f_x){
      count <- count+1
      samples[count] <- sample
    }
   
  }
  return(samples)

}
set.seed(12345)
x <- rejection_sample(10000)
#hist(x,breaks = 100)
data <- data.frame(x)
ggplot(data,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using  rejection sampling", x = "x", y ="Frequency")+ theme_minimal()




#b
com_sample <- function(n){
  U1 <- runif(n/2) #50%
  Y1 <- 1-sqrt(1-U1)
  
  U2 <- runif(n/2)
  Y2 <- -(1-sqrt(1-U2))
  
  X <- c(Y1,Y2)
  return(X)
  
}
set.seed(12345)
composition_samples <- com_sample(10000)
#hist(composition_samples,breaks = 100)
dataCS <- data.frame(composition_samples)
ggplot(dataCS,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using composition sampling", x = "x", y ="Frequency")+ theme_minimal()

#c.
set.seed(12345)
U1 <- runif(10000,0,1)
U2 <- runif(10000,0,1)

U3 <- U1-U2
#hist(U3,breaks = 100)
dataDifference <- data.frame(U3)
ggplot(dataDifference,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using the difference of uniform distributions", x = "x", y ="Frequency")+ theme_minimal()
#d
cat("Variance of X: ",var(U3),"\n")
```

# QUESTION 1: Sampling algorithms for a triangle distribution

To generate draws of a random variable X with this function: 
$$f(x) =
\begin{cases}
  0, & x < -1 \text{ or } x > 1, \\
  x + 1, & -1 \leq x \leq 0, \\
  1 - x, & 0 < x \leq 1.
\end{cases}$$
First, we implemented `rejection sampling method`. We choose uniform distribution as the envelope function $e(x)$:
$$
e(x) = 
\begin{cases} 
1 & \text{if } -1 \leq x \leq 1, \\
0 & \text{otherwise}.
\end{cases}
$$

This function always takes the value 1 in the interval[-1,1],which is greater than or eaqual to $f(x)$ for all $x$. Then, we implemented the rejection sampling algorithm to generate 10,000 random variables. The histogram is shown as follows:

```{r,echo=FALSE, fig.dim=c(4, 2.5), fig.align='center' }
data <- data.frame(x)
ggplot(data,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using  rejection sampling", x = "x", y ="Frequency")+ theme_minimal()

```

Then, we implemented composition sampling method to generate draws. Let $Y$ be a random variable following this density function:
$$y=
\begin{cases}
  2 - 2x, & 0 \leq x \leq 1, \\
  0, &otherwise.
\end{cases}$$
$-Y$ has the similar triangle distribution in the interval [-1,0]. The graph of $-Y$ and $Y$ is symmetric about the y-axis. By using `Inverse CDF method`, we generated $Y$ and by the property of symmetry, we could generate $-Y$. The areas under the graphs of $Y$ and $-Y$ are the same, therefore the mixing parameters are both 0.5. Then we could generate 10,000 random variables and the histogram is shown as follows:

```{r,echo=FALSE, fig.dim=c(4, 2.5), fig.align='center'}
composition_samples <- com_sample(10000)
dataCS <- data.frame(composition_samples)
ggplot(dataCS,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using composition sampling", x = "x", y ="Frequency")+ theme_minimal()

```

When $U1$, $U2$ are two independent `Unif[0,1]` distributed random variables, $U1 - U2$
 has the same distribution as X. We used this result and generated random variables and the histogram is shown as follows:
 
```{r,echo=FALSE, fig.dim=c(4, 2.5), fig.align='center'}
U3 <- U1-U2
dataDifference <- data.frame(U3)
ggplot(dataDifference,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using the difference of uniform distributions", x = "x", y ="Frequency")+ theme_minimal()

```
 
**Comparison of Three Sampling Methods**  

`Rejection Sampling Method` :  
1.The amount of programming:    
Requires generating  random variables for the envelope function and then generating a random number in [0,1], followed by accepting or rejecting samples based on the density function.  
2. The number of random value generation and other time consuming:  
Two uniform samples are generated for each sample, but samples may be rejected, leading to more random value generation. Additionally, the choice of the envelope function can affect the acceptance rate, resulting in more time-consuming operation.

`Composition Sampling Method`:  
1.The amount of programming:  
Requires generating Y using ` Inverse CDF method`,we should calculate the inverse CDF of Y and then use 
symmetry to obtain $-Y$ and finally obtain $X$.  
2. The number of random value generation and other time consuming:  
Efficient. Only one random variable generation is needed for each sample.Calculating the inverse CDF may lead to additional time consumption.However, it is still more efficient compared to the `Composition Sampling Method`,because it doesn't require generating waste samples.  

`Difference of Uniforms`:  
1.The amount of programming:   
Most efficient. It only requires generating random variables from two uniform distributions and taking their difference.  
2. The number of random value generation and other time consuming:  
Only two uniform samples generated for each sample.And no need for rejection setps and other time consuming steps.

Based on the analysis above, we choose the third method _`Difference of Uniforms`_ when generating samples of $X$. The variance of $X$ is :
```{r , echo=FALSE}
cat("Variance of X: ",var(U3),"\n")
```

# Appendix
## Question 1
```{r eval=FALSE}
#a
library(ggplot2)
fx <- function(x){
  if(x< -1|x > 1){
    return(0)}
  if(-1 <= x&& x <= 0){
    return(x+1)}
  if(0 < x && x <= 1){
    return(1-x)}
}

rejection_sample <- function(n){
  samples <- numeric(n)
  count <- 0
  
  while (count<n) {
    sample <- runif(1,-1,1) #sampling from uniform distribution
    U <- runif(1)
    f_x <- fx(sample)
    
    if(U <= f_x){
      count <- count+1
      samples[count] <- sample
    }
   
  }
  return(samples)

}
set.seed(12345)
x <- rejection_sample(10000)
#hist(x,breaks = 100)
data <- data.frame(x)
ggplot(data,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using  rejection sampling", x = "x", y ="Frequency")+ theme_minimal()




#b
com_sample <- function(n){
  U1 <- runif(n/2) #50%
  Y1 <- 1-sqrt(1-U1)
  
  U2 <- runif(n/2)
  Y2 <- -(1-sqrt(1-U2))
  
  X <- c(Y1,Y2)
  return(X)
  
}
set.seed(12345)
composition_samples <- com_sample(10000)
#hist(composition_samples,breaks = 100)
dataCS <- data.frame(composition_samples)
ggplot(dataCS,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using composition sampling", x = "x", y ="Frequency")+ theme_minimal()

#c.
set.seed(12345)
U1 <- runif(10000,0,1)
U2 <- runif(10000,0,1)

U3 <- U1-U2
#hist(U3,breaks = 100)
dataDifference <- data.frame(U3)
ggplot(dataDifference,aes(x=x))+
  geom_histogram(bins = 100,fill="#4E79A7",color="black")+
labs(title = "Histogram of x using the difference of uniform distributions", x = "x", y ="Frequency")+ theme_minimal()

#d
cat("Variance of X: ",var(U3),"\n")
```

