---
title: "Lab 6 - Computational Statistics (732A89)"
author: "Helena Llorens Lluís (hllor282), Yi Yang (yiyan338)"
output: pdf_document
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim=c(4, 2.5), fig.align='center', message = F, warning = F)
```

```{r, echo=FALSE}
library(ggplot2)
library(knitr)
```


# QUESTION 1: EM algorithm

Expectation-Maximization (EM) is an iterative optimization algorithm used for parameter estimation in probabilistic models with latent variables. In this report, we modify the provided EM algorithm to handle a mixture of three normal distributions.

Starting from the provided two-component EM algorithm, we introduced the following changes:

- Increased the number of mixture components from two to three.

- Adjusted the initialization of parameters accordingly.

- Modified the Expectation (E) step to compute probabilities for three components.

- Updated the Maximization (M) step to re-estimate parameters based on three components.

- Revised the stopping criterion to ensure scale-invariance by normalizing the convergence measure.

```{r, echo=FALSE}
# EM algorithm
em <- function(data, eps = 0.0001){
  n      <- length(data)
  pi     <- matrix(NA, ncol = 3, nrow = n)  
  
  # Random initialization of the parameters
  p <- rep(1/3, 3)
  sigma <- rep(sd(data)*2/3, 3)
  mu <- c(mean(data) - sigma[1]/2, mean(data), mean(data) + sigma[1]/2)
  pv <- c(p, mu, sigma)
  cc  <- eps + 100
  mu_list <- c(mu)
  sigma_list <- c(sigma)
  
  while (cc > eps){
    # save previous parameter vector
    pv1  <- pv                 
    
    # E step 
    for (j in 1:n){
      pi1 <- p[1]*dnorm(data[j], mean=mu[1], sd=sigma[1])
      pi2 <- p[2]*dnorm(data[j], mean=mu[2], sd=sigma[2])
      pi3 <- p[3]*dnorm(data[j], mean=mu[3], sd=sigma[3])
      
      total <- pi1+pi2+pi3
      pi[j,] <- c(pi1/total, pi2/total, pi3/total)
    }

    # M step 
    p <- colSums(pi)/n
    mu[1] <- sum(pi[, 1]*data)/(p[1]*n)
    mu[2] <- sum(pi[, 2]*data)/(p[2]*n)
    mu[3] <- sum(pi[, 3]*data)/(p[3]*n)
    sigma[1] <- sqrt(sum(pi[, 1]*((data-mu[1])^2))/(p[1]*n))
    sigma[2] <- sqrt(sum(pi[, 2]*((data-mu[2])^2))/(p[2]*n))
    sigma[3] <- sqrt(sum(pi[, 3]*((data-mu[3])^2))/(p[3]*n))
    
    # stopping criteria
    pv <- c(p, mu, sigma)
    cc     <- sum((pv - pv1)^2) / sum(pv1^2)  # a convergence criterion, maybe not the best one
    mu_list <- rbind(mu_list, mu)
    sigma_list <- rbind(sigma_list, sigma)
    
  }
  res <- list(final_probs = pv[1:3], 
              final_mus = pv[4:6],
              final_sigmas = pv[7:9],
              mu = mu_list, 
              sigma = sigma_list)
  return(res)
}
```


After loading the `dat3p` dataset, we plot a histogram to visualize its distribution.

```{r, echo=FALSE}
load("threepops.Rdata")
data <- dat3p
df <- data.frame(data)
ggplot(df, aes(x = df[, 1])) + 
  geom_histogram(col = "black", fill = "#4E79A7") + 
  labs(title = "Histogram of dat3p", x = "Value", y = "Frequency") +
  theme_minimal()
```

The data exhibits multimodal behavior, suggesting the presence of multiple underlying distributions.

Then, we applied the EM algorithm with a stopping threshold of eps = 0.000001. The final estimated parameters of the normal mixture model are showed on the following table.

```{r, echo=FALSE}
res <- em(data)
final_probs <- res$final_probs
final_mus <- res$final_mus
final_sigmas <- res$final_sigmas

result_table <- data.frame(final_probs, final_mus, final_sigmas)
colnames(result_table) <- c("p", "mu", "sigma")
rownames(result_table) <- c("Component 1", "Component 2", "Component 3")
kable(result_table, 
      caption = "Final estimated parameters of the normal mixture model")
```

These values suggest that the algorithm successfully identified three distinct normal distributions underlying the dataset.

To assess convergence, we plotted the estimated $\mu$ and $\sigma$ values across iterations.

```{r, echo=FALSE, fig.cap='Plot of mu vs iteration number'}
#  plots of current estimates for each model parameter versus the iteration-number

# mu plot
mus <- data.frame(res$mu)
ggplot(data = mus, aes(x = 1:nrow(mus))) + 
  geom_line(aes(y = mus[, 1]), col = "#4E79A7", size = 0.7) + 
  geom_line(aes(y = mus[, 2]), col = "#F28E2B", size = 0.7) + 
  geom_line(aes(y = mus[, 3]), col = "#E15759", size = 0.7) + 
  theme_minimal() + 
  labs(x = "Iterations", y = expression(mu))
```

```{r, echo=FALSE, fig.cap='Plot of sigma vs iteration number'}
# sigma plot
sigmas <- data.frame(res$sigma)
ggplot(data = sigmas, aes(x = 1:nrow(sigmas))) + 
  geom_line(aes(y = sigmas[, 1]), col = "#4E79A7", size = 0.7) + 
  geom_line(aes(y = sigmas[, 2]), col = "#F28E2B", size = 0.7) + 
  geom_line(aes(y = sigmas[, 3]), col = "#E15759", size = 0.7) + 
  theme_minimal() + 
  labs(x = "Iterations", y = expression(sigma))
```

Both estimates stabilize after a certain number of iterations, supporting convergence.

The EM algorithm successfully fitted a three-component normal mixture model. The convergence plots validate that the algorithm converged for each parameter. Additionally, modifying the stopping criterion ensured robustness against data scaling.


# Question 2 Simulated annealing

Fisrt, we plot the data and select randomly 22 clients.The red points show the initial cilents that we have choosen, which will be used in the simulated algorithm.
```{r echo=FALSE,fig.cap='Plot of data', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
library(ggplot2)
load("bankdata.Rdata")
data <- as.data.frame(bankdata)
nclients <- dim(data)[1]  # number of individuals in the dataset
#a
set.seed(12345)
initial <- sample(1:nclients,22)
df <- data.frame(age=data$age,balance=data$balance)
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  geom_point(data=df[initial,],aes(age,balance),color="red",size=1)+
  theme_minimal()
```
We now implement `simulated annealing` algorithm to minize the `criterion function`.To ensure we always have subsets with 22 clients, we **remove one client from the subsets and randomly select another client from the whole dataset in each iteration**.
```{r echo=FALSE }
crit <- function(dat, subs){
  s <- length(subs)
  dist <- matrix(rep(NA, nclients*s), ncol=s)
  for (i in 1:s){
    dist[, i] <- sqrt((dat[,1]-dat[subs[i],1])^2+(dat[,2]-dat[subs[i],2])^2)
  }
  sum(apply(dist, 1, min))
}

simulated_annealing <- function(data,initial,max_stage,initial_temperature,alpha,initial_iteration,beta){
  
  current_subsample <- initial
  current_crit <- crit(data,current_subsample)
  best_subsample <- current_subsample
  best_crit <- current_crit
  
  temperature <- initial_temperature
  iter_per_stage <- initial_iteration
  crit_values <- numeric()
  total_iter <- 0
  
  for(stage in 1:max_stage){
    
    for (i in 1:iter_per_stage) {
      
      total_iter <- total_iter + 1
      #generate new subsample
      new_subsample <- current_subsample
      remove_index <- sample(1: length(current_subsample),1)
      #select one sample from dataset
      add_index <- sample(setdiff(1:nclients,current_subsample),1)
      new_subsample[remove_index] <- add_index
      
      #calculate new crit
      new_crit <- crit(data,new_subsample)
      
      #calculate the acceptance probability
      differernce_crit <- new_crit- current_crit
      accept_probability <- ifelse(differernce_crit<0,1,exp(-differernce_crit/temperature))
      
      if(runif(1)<accept_probability){
        current_subsample <- new_subsample
        current_crit <- new_crit
        
      }
      
      if(current_crit<best_crit){
        best_subsample <- current_subsample
        best_crit <- current_crit
      }
    
      crit_values[total_iter] <- current_crit
    }
    
      temperature <- temperature*alpha
      iter_per_stage <- iter_per_stage*beta
      

  }
  
  return(list(
    best_subsample=best_subsample,
    best_crit=best_crit,
    crit_values=crit_values
  ))
  
  
}
```

We then try different combinations of schedules, temperatures and iterations to find a better solution.The fist combination is :$$initial temperature=200$$  $$max stage = 10$$ $$\alpha=0.9$$  $$\beta=1.5$$ $$initial iteration=100$$ 
The plots of the criterion-value versus the iteration number ,as well as the final clients marked are shown:
```{r echo=FALSE,fig.cap='Criterion-value Versus the Iteration Number', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
set.seed(12345)
result1 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=100,beta=1.5)
#plot(result1$crit_values,type = "l",col="red",xlab="Iteration Number",ylab="Criterition Value",main="Criterion-Value versus Iteration")

dfcritvalue1 <- data.frame(IterationNumber=1:length(result1$crit_values),
                           CriterionValue=result1$crit_values)
ggplot(dfcritvalue1,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()


```

```{r,echo=FALSE,fig.cap='Final subset', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
dfresult1 <- data[result1$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult1,aes(age,balance),color="blue",size=1)+
  theme_minimal()
```

The best crit value is:
```{r echo=FALSE }
print(result1$best_crit)
```


Then we try the second combination:$$initial temperature=200$$  $$max stage = 10$$ $$\alpha=0.9$$  $$\beta=1.5$$ $$initial iteration=10$$
```{r echo=FALSE,fig.cap='Criterion-value Versus the Iteration Number', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
set.seed(12345)
result2 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=10,beta=1.5)

dfcritvalue2 <- data.frame(IterationNumber=1:length(result2$crit_values),
                           CriterionValue=result2$crit_values)
ggplot(dfcritvalue2,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()
```

```{r echo=FALSE,fig.cap='Final subset', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
dfresult2 <- data[result2$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult2,aes(age,balance),color="blue",size=1)+
  theme_minimal()

```

The best crit value is:
```{r echo=FALSE }
print(result2$best_crit)
```

**Comparison**  
1.Solution quality:  
The  second combination achieved a lower criterion value (15778.63) compared to the first combination (15815.4).  
2.Convergence Speed:  
The first combination required a significantly higher number of iterations (9000) to reach the global minimum, indicating a slower convergence.  
The second combination reached its best value much faster, around 370 iterations, indicating a quicker convergence.  
3.Parameter Impact:  
The higher initial iteration count (100) in the first combination allowed for more extensive exploration in each stage, which likely contributed to finding a better solution but at the cost of more iterations.   
The lower initial iteration count (10) in the second combination led to faster convergence, allowing the algorithm to quickly find a better solution.

**Conclusion**  
**The second combination** outperforms the first combination in both solution quality and convergence speed. This suggests that a lower initial iteration count with a higher iteration increase factor can lead to quicker and better solutions in this context.


# APPENDIX
## Question 1
```{r, eval=FALSE}
library(ggplot2)
library(knitr)

# EM algorithm
em <- function(data, eps = 0.0001){
  n      <- length(data)
  pi     <- matrix(NA, ncol = 3, nrow = n)  
  
  # Random initialization of the parameters
  p <- rep(1/3, 3)
  sigma <- rep(sd(data)*2/3, 3)
  mu <- c(mean(data) - sigma[1]/2, mean(data), mean(data) + sigma[1]/2)
  pv <- c(p, mu, sigma)
  cc  <- eps + 100
  mu_list <- c(mu)
  sigma_list <- c(sigma)
  
  while (cc > eps){
    # save previous parameter vector
    pv1  <- pv                 
    
    # E step 
    for (j in 1:n){
      pi1 <- p[1]*dnorm(data[j], mean=mu[1], sd=sigma[1])
      pi2 <- p[2]*dnorm(data[j], mean=mu[2], sd=sigma[2])
      pi3 <- p[3]*dnorm(data[j], mean=mu[3], sd=sigma[3])
      
      total <- pi1+pi2+pi3
      pi[j,] <- c(pi1/total, pi2/total, pi3/total)
    }

    # M step 
    p <- colSums(pi)/n
    mu[1] <- sum(pi[, 1]*data)/(p[1]*n)
    mu[2] <- sum(pi[, 2]*data)/(p[2]*n)
    mu[3] <- sum(pi[, 3]*data)/(p[3]*n)
    sigma[1] <- sqrt(sum(pi[, 1]*((data-mu[1])^2))/(p[1]*n))
    sigma[2] <- sqrt(sum(pi[, 2]*((data-mu[2])^2))/(p[2]*n))
    sigma[3] <- sqrt(sum(pi[, 3]*((data-mu[3])^2))/(p[3]*n))
    
    # stopping criteria
    pv <- c(p, mu, sigma)
    cc     <- sum((pv - pv1)^2) / sum(pv1^2)  # a convergence criterion, maybe not the best one
    mu_list <- rbind(mu_list, mu)
    sigma_list <- rbind(sigma_list, sigma)
    
  }
  res <- list(final_probs = pv[1:3], 
              final_mus = pv[4:6],
              final_sigmas = pv[7:9],
              mu = mu_list, 
              sigma = sigma_list)
  return(res)
}

# load data
load(file = "D:\\liu\\CompStat\\Computational-Statistics\\lab6\\threepops.Rdata")
data <- dat3p

# Histogram of the data
df <- data.frame(data)
ggplot(df, aes(x = df[, 1])) + 
  geom_histogram(col = "black", fill = "#4E79A7") + 
  labs(title = "Histogram of dat3p", x = "Value", y = "Frequency") +
  theme_minimal()

# Run algorithm on the data
res <- em(data)
final_probs <- res$final_probs
final_mus <- res$final_mus
final_sigmas <- res$final_sigmas

# results table
result_table <- data.frame(final_probs, final_mus, final_sigmas)
colnames(result_table) <- c("p", "mu", "sigma")
rownames(result_table) <- c("Component 1", "Component 2", "Component 3")
kable(result_table, 
      caption = "Final estimated parameters of the normal mixture model")

#  plots of current estimates for each model parameter versus the iteration-number
# mu plot
mus <- data.frame(res$mu)
ggplot(data = mus, aes(x = 1:nrow(mus))) + 
  geom_line(aes(y = mus[, 1]), col = "#4E79A7", size = 0.7) + 
  geom_line(aes(y = mus[, 2]), col = "#F28E2B", size = 0.7) + 
  geom_line(aes(y = mus[, 3]), col = "#E15759", size = 0.7) + 
  theme_minimal() + 
  labs(x = "Iterations", y = expression(mu))

# sigma plot
sigmas <- data.frame(res$sigma)
ggplot(data = sigmas, aes(x = 1:nrow(sigmas))) + 
  geom_line(aes(y = sigmas[, 1]), col = "#4E79A7", size = 0.7) + 
  geom_line(aes(y = sigmas[, 2]), col = "#F28E2B", size = 0.7) + 
  geom_line(aes(y = sigmas[, 3]), col = "#E15759", size = 0.7) + 
  theme_minimal() + 
  labs(x = "Iterations", y = expression(sigma))
```

## Question 2
```{r, eval=FALSE}
  #question2
library(ggplot2)
load("bankdata.Rdata")
data <- as.data.frame(bankdata)
nclients <- dim(data)[1]  # number of individuals in the dataset, here 4364


crit <- function(dat, subs){
  s <- length(subs)
  dist <- matrix(rep(NA, nclients*s), ncol=s)
  for (i in 1:s){
    dist[, i] <- sqrt((dat[,1]-dat[subs[i],1])^2+(dat[,2]-dat[subs[i],2])^2)
  }
  sum(apply(dist, 1, min))
}


#a
set.seed(12345)
initial <- sample(1:nclients,22)
df <- data.frame(age=data$age,balance=data$balance)
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=2,alpha=0.6)+
  geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  theme_minimal()

#b simulated annealing 



exchange_clients <- function(initial,nclients){
  
  remove_index <- sample(1:length(initial),1)
  #select one sample from dataset
  add_index <- sample(setdiff(1:nclients,initial),1)
  new_initial <- initial
  new_initial[remove_index] <- add_index
  return(new_initial)
  
}

simulated_annealing <- function(data,initial,max_stage,initial_temperature,alpha,initial_iteration,beta){
  
  current_subsample <- initial
  current_crit <- crit(data,current_subsample)
  best_subsample <- current_subsample
  best_crit <- current_crit
  
  temperature <- initial_temperature
  iter_per_stage <- initial_iteration
  crit_values <- numeric()
  total_iter <- 0
  
  for(stage in 1:max_stage){
    
    for (i in 1:iter_per_stage) {
      
      total_iter <- total_iter + 1
      #generate new subsample
      new_subsample <- current_subsample
      remove_index <- sample(1: length(current_subsample),1)
      #select one sample from dataset
      add_index <- sample(setdiff(1:nclients,current_subsample),1)
      new_subsample[remove_index] <- add_index
      
      #calculate new crit
      new_crit <- crit(data,new_subsample)
      
      #calculate the acceptance probability
      differernce_crit <- new_crit- current_crit
      accept_probability <- ifelse(differernce_crit<0,1,exp(-differernce_crit/temperature))
      
      if(runif(1)<accept_probability){
        current_subsample <- new_subsample
        current_crit <- new_crit
        
      }
      
      if(current_crit<best_crit){
        best_subsample <- current_subsample
        best_crit <- current_crit
      }
    
      crit_values[total_iter] <- current_crit
    }
    
      temperature <- temperature*alpha
      iter_per_stage <- iter_per_stage*beta
      

  }
  
  return(list(
    best_subsample=best_subsample,
    best_crit=best_crit,
    crit_values=crit_values
  ))
  
  
}


set.seed(12345)
result1 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=100,beta=1.5)
#plot(result1$crit_values,type = "l",col="red",xlab="Iteration Number",ylab="Criterition Value",main="Criterion-Value versus Iteration")

dfcritvalue1 <- data.frame(IterationNumber=1:length(result1$crit_values),
                           CriterionValue=result1$crit_values)
ggplot(dfcritvalue1,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()

dfresult1 <- data[result1$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=2,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult1,aes(age,balance),color="blue",size=2)+
  theme_minimal()

print(result1$best_crit)

set.seed(12345)
result2 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=10,beta=1.5)

dfcritvalue2 <- data.frame(IterationNumber=1:length(result2$crit_values),
                           CriterionValue=result2$crit_values)
ggplot(dfcritvalue2,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()

dfresult2 <- data[result2$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=2,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult2,aes(age,balance),color="blue",size=2)+
  theme_minimal()
print(result2$best_crit)


```

