---
title: "question2"
author: "Helena Llorens Lluís (hllor282), Yi Yang (yiyan338)"
date: "2025-03-03"
output: pdf_document
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```


```{r echo=FALSE}
library(ggplot2)
library(knitr)
```


# Question 2 Simulated annealing

Fisrt, we plot the data and select randomly 22 clients.The red points show the initial cilents that we have choosen, which will be used in the simulated algorithm.
```{r echo=FALSE,fig.cap='Plot of data', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
library(ggplot2)
load("bankdata.Rdata")
data <- as.data.frame(bankdata)
nclients <- dim(data)[1]  # number of individuals in the dataset
#a
set.seed(12345)
initial <- sample(1:nclients,22)
df <- data.frame(age=data$age,balance=data$balance)
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  geom_point(data=df[initial,],aes(age,balance),color="red",size=1)+
  theme_minimal()
```
We now implement `simulated annealing` algorithm to minize the `criterion function`.To ensure we always have subsets with 22 clients, we **remove one client from the subsets and randomly select another client from the whole dataset in each iteration**.
```{r echo=FALSE }
crit <- function(dat, subs){
  s <- length(subs)
  dist <- matrix(rep(NA, nclients*s), ncol=s)
  for (i in 1:s){
    dist[, i] <- sqrt((dat[,1]-dat[subs[i],1])^2+(dat[,2]-dat[subs[i],2])^2)
  }
  sum(apply(dist, 1, min))
}

simulated_annealing <- function(data,initial,max_stage,initial_temperature,alpha,initial_iteration,beta){
  
  current_subsample <- initial
  current_crit <- crit(data,current_subsample)
  best_subsample <- current_subsample
  best_crit <- current_crit
  
  temperature <- initial_temperature
  iter_per_stage <- initial_iteration
  crit_values <- numeric()
  total_iter <- 0
  
  for(stage in 1:max_stage){
    
    for (i in 1:iter_per_stage) {
      
      total_iter <- total_iter + 1
      #generate new subsample
      new_subsample <- current_subsample
      remove_index <- sample(1: length(current_subsample),1)
      #select one sample from dataset
      add_index <- sample(setdiff(1:nclients,current_subsample),1)
      new_subsample[remove_index] <- add_index
      
      #calculate new crit
      new_crit <- crit(data,new_subsample)
      
      #calculate the acceptance probability
      differernce_crit <- new_crit- current_crit
      accept_probability <- ifelse(differernce_crit<0,1,exp(-differernce_crit/temperature))
      
      if(runif(1)<accept_probability){
        current_subsample <- new_subsample
        current_crit <- new_crit
        
      }
      
      if(current_crit<best_crit){
        best_subsample <- current_subsample
        best_crit <- current_crit
      }
    
      crit_values[total_iter] <- current_crit
    }
    
      temperature <- temperature*alpha
      iter_per_stage <- iter_per_stage*beta
      

  }
  
  return(list(
    best_subsample=best_subsample,
    best_crit=best_crit,
    crit_values=crit_values
  ))
  
  
}
```

We then try different combinations of schedules, temperatures and iterations to find a better solution.The fist combination is :$$initial temperature=200$$  $$max stage = 10$$ $$\alpha=0.9$$  $$\beta=1.5$$ $$initial iteration=100$$ 
The plots of the criterion-value versus the iteration number ,as well as the final clients marked are shown:
```{r echo=FALSE,fig.cap='Criterion-value Versus the Iteration Number', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
set.seed(12345)
result1 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=100,beta=1.5)
#plot(result1$crit_values,type = "l",col="red",xlab="Iteration Number",ylab="Criterition Value",main="Criterion-Value versus Iteration")

dfcritvalue1 <- data.frame(IterationNumber=1:length(result1$crit_values),
                           CriterionValue=result1$crit_values)
ggplot(dfcritvalue1,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()


```

```{r,echo=FALSE,fig.cap='Final subset', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
dfresult1 <- data[result1$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult1,aes(age,balance),color="blue",size=1)+
  theme_minimal()
```

The best crit value is:
```{r echo=FALSE }
print(result1$best_crit)
```


Then we try the second combination:$$initial temperature=200$$  $$max stage = 10$$ $$\alpha=0.9$$  $$\beta=1.5$$ $$initial iteration=10$$
```{r echo=FALSE,fig.cap='Criterion-value Versus the Iteration Number', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
set.seed(12345)
result2 <- simulated_annealing(data=data,initial=initial,max_stage=10,initial_temperature=200,alpha=0.9,initial_iteration=10,beta=1.5)

dfcritvalue2 <- data.frame(IterationNumber=1:length(result2$crit_values),
                           CriterionValue=result2$crit_values)
ggplot(dfcritvalue2,aes(x=IterationNumber,y=CriterionValue))+
  geom_line(color="#4E79A7")+
  labs(x="IterationNumber",
       y="CriterionValue",
       title = "Criterion-Value versus Iteration")+
  theme_minimal()
```

```{r echo=FALSE,fig.cap='Final subset', fig.dim=c(4, 2.5), fig.align='center', fig.pos="H"}
dfresult2 <- data[result2$best_subsample, ]
ggplot(df,aes(age,balance))+
  geom_point(color="#4E79A7",size=1,alpha=0.6)+
  #geom_point(data=df[initial,],aes(age,balance),color="red",size=2)+
  geom_point(data=dfresult2,aes(age,balance),color="blue",size=1)+
  theme_minimal()

```

The best crit value is:
```{r echo=FALSE }
print(result2$best_crit)
```
**Comparison**  
1.Solution quality:  
The  second combination achieved a lower criterion value (15778.63) compared to the first combination (15815.4).  
2.Convergence Speed:  
The first combination required a significantly higher number of iterations (9000) to reach the global minimum, indicating a slower convergence.  
The second combination reached its best value much faster, around 370 iterations, indicating a quicker convergence.  
3.Parameter Impact:  
The higher initial iteration count (100) in the first combination allowed for more extensive exploration in each stage, which likely contributed to finding a better solution but at the cost of more iterations.   
The lower initial iteration count (10) in the second combination led to faster convergence, allowing the algorithm to quickly find a better solution.

**Conclusion**  
**The second combination** outperforms the first combination in both solution quality and convergence speed. This suggests that a lower initial iteration count with a higher iteration increase factor can lead to quicker and better solutions in this context.